---
title: 'STAT 420: Homework 02'
author: "Spring 2020, Yu Wu"
date: 'Due: Tuesday, February 11 by 11:30 PM CT'
output:
  html_document:
    theme: readable
    toc: yes
  pdf_document:
    toc: yes
---

# Assignment

## Exercise 1 (Writing Simple Functions)

For each of the following parts, use the following vectors:

```{r}
a = 1:10
b = 10:1
c = rep(1, times = 10)
d = 2 ^ (1:10)
```

**(a)** Write a function called `sum_of_squares`.

- Arguments:
    - A vector of numeric data `x`.
- Output:
    - The sum of the squares of the elements of the vector. $\sum_{i = 1}^n x_i^2$
    
Provide your function, as well as the result of running the following code:

```{r, eval = FALSE}
sum_of_squares <- function(x){
  sum(x^2)
}
sum_of_squares(x = a)
sum_of_squares(x = c(c, d))
```

**(b)** Write a function called `sum_of_power`.

- Arguments:
    - A vector of numeric data `x`.
    - `p` which should have the default value of `2`.
- Output:
    - $\sum_{i = 1}^n x_i^{p}$
    
Provide your function, as well as the result of running the following code:

```{r, eval = FALSE}
sum_of_power <- function(x, p=2){
  sp <- 0
  for (i in x) {
    sp <- sp + i^p
  }
  return(sp)
}

sum_of_power(x = a)
sum_of_power(x = a, p = 3)
sum_of_power(x = a, p = a)
sum_of_power(x = a, p = c(1, 2))
```

**(c)** Write a function called `rms_diff`.

- Arguments:
    - A vector of numeric data `x`.
    - A vector of numeric data `y`.
- Output:
    - $\sqrt{\frac{1}{n}\sum_{i = 1}^n (x_i - y_i)^{2}}$
    
Provide your function, as well as the result of running the following code:

```{r, eval = FALSE}
rms_diff <- function(x,y){
  sum(x-y)^2/length(x)
}

rms_diff(x = a, y = b)
rms_diff(x = d, y = c)
rms_diff(x = d, y = 1)
rms_diff(x = a, y = 0) ^ 2 * length(a)
```

## Exercise 2 (Plotting, Testing)

For this exercise we will use [the data that is stored in `intelligence.csv`](intelligence.csv) which records IQs of a random sample of residents of Pawnee and Eagleton, Indiana.

**(a)** Load the data from `intelligence.csv` into a variable in `R` called `intelligence`. Show the code used to do this. 

```{r}
intelligence <- read.csv('intelligence.csv')
```

**(b)** Create a side-by-side boxplot that compares the IQs across the two towns. Be sure to give the plot a title and label the axes appropriately.

```{r}
boxplot(iq ~ town, data = intelligence, main = "Side-by-side Box Plot of IQs Across the Two Towns")
```

**(c)** Are people from Eagleton smarter than people from Pawnee? Perform an appropriate statistical test using the given sample data. That is, test $H_0: \mu_E = \mu_P$ vs $H_1: \mu_E > \mu_P$, where

- $\mu_E$ is the mean IQ of a resident of Eagleton.
- $\mu_P$ is the mean IQ of a resident of Pawnee.

```{r}
t.test(iq~town, data = intelligence, alternative = 'greater')
```


Explicitly state the p-value of the test and the resulting statistical decision at a significance level $\alpha = 0.10$. Interpret the results in the context of the problem.

**(d)** Repeat **(c)** using a two-sided alternative hypothesis. What changes?

```{r}
t.test(iq~town, data = intelligence)
```

The p-value doubles.

## Exercise 3 (Writing More Functions)

In this exercise we will write our own functions related to performing a one-sample $t$ test. That is $H_{0}: \mu = \mu_{0}$ versus $H_{1}: \mu \neq \mu_{0},$ where $\mu_{0}$ is the hypothesized value of $\mu$.

Throughout this exercise you may **not** use the `t.test()` function inside your functions. You may use it to check your work separately, but no such double-checks should appear in your final report.

Some built in `R` functions that may be useful to you when writing your functions include: `c()`, `ifelse()`, `mean()`, `sd()`, `abs()`, `length()`, `sqrt()`, and `pt()`.

**(a)** Write a function called `do_t_test` which takes two inputs:

- `x`: A vector which stores observations.
- `mu`: The hypothesized value of $\mu$ which defaults to `0`.

The function should output:

- The value of the test statistic, $t$.
- The p-value of the test. The function only needs to be able to handle a two-sided alternative.

In order to output both, consider using `c(t, pval)` as the last line of your function, and store those two values elsewhere in the body of your function.

```{r}
do_t_test <- function(x, mu = 0){
  n <- length(x)
  x_bar <- sum(x)/n
  t <- (x_bar-mu)/(sd(x)/sqrt(n))
  pv <- 2*pt(-abs(t),df=n-1)
  return(c(t, pv))
}
```

**(b)** Write a function called `make_decision` which takes two inputs:

- `pval`: The p-value of a test.
- `alpha`: The significance level of a test. Set a default value of `0.05`.

The function should output `"Reject!"` or `"Fail to Reject."` based on the comparison of `pval` to `alpha`.


```{r}
make_decision <- function(pval, alpha = 0.05){
  if (pval < alpha/2){
    return("Fail to Reject")
  }
  return("Reject")
}
```

**(c)** Now we will test the quality of your functions from parts **(a)** and **(b)**. Run the following code:

```{r, eval = FALSE}
set.seed(114)
y = rnorm(25, 1.4, 1)
pval = do_t_test(y, mu = 2)[2]
pval
make_decision(pval, alpha = 0.10)


```

If your `do_t_test()` and `make_decision()` functions are correct, you should obtain a decision of `"Fail to Reject."` You will also be evaluated on whether the numeric p-value you obtain is correct.





## Exercise 4 (CLT Simulation)

For this exercise we will simulate from the exponential distribution. If a random variable $X$ has an exponential distribution with rate parameter $\lambda$, the pdf of $X$ can be written

\[
f(x;\lambda) = \lambda e ^ {-\lambda x}
\]

for $x \geq 0$.

Also recall,

\[
\begin{aligned}
\mu &= E[X] = \frac{1}{\lambda} \\
\sigma^ 2 &= Var[X] = \frac{1}{\lambda ^ 2}
\end{aligned}
\]

**(a)** This exercise relies heavily on generating random observations. To make this reproducible we will set a seed for the randomization. Alter the following code to make `birthday` store your birthday in the format: `yyyymmdd`. For example, [William Gosset](https://en.wikipedia.org/wiki/William_Sealy_Gosset), better known as *Student*, was born on June 13, 1876, so he would use:

```{r}
birthday = 19981121
set.seed(birthday)
```

**(b)** Simulate 10000 samples of size **5** from an exponential distribution with $\lambda = 2$. Store the mean of each sample in a vector. Plot a histogram of these sample means. (Be sure to give it a title, and label the axes appropriately.) Based on the histogram, do you think the central limit theorem applies here?

```{r}
means <- numeric(10000)
for (i in 1:10000) {
  means[i] <- mean(rexp(5, 2))
}
hist(means, main = "Histogram of Sample Means", xlab = "Sample Means")
```

It does not apply here because the size of each sample is only 5 and it is not big enough.

**(c)** Simulate 10000 samples of size **100** from an exponential distribution with $\lambda = 2$. Store the mean of each sample in a vector. Plot a histogram of these sample means. (Be sure to give it a title, and label the axes appropriately.) Based on the histogram, do you think the central limit theorem applies here?

```{r}
means <- numeric(10000)
for (i in 1:10000) {
  means[i] <- mean(rexp(100, 2))
}
hist(means, main = "Histogram of Sample Means", xlab = "Sample Means")
```

It applies here because the sample size is now 100 and it is big enough to show normal distribution.

**(d)** We just repeated ourselves, so that means we probably should be writing a function. Write a function called `sim_xbars_exp` which takes three inputs:

- The number of samples to simulate.
- The sample size.
- The rate parameter of an exponential distribution.

The function should output a vector of sample means which are the result of sampling from an exponential distribution as specified by the inputs.

Use your function to simulate 25000 samples of size **50** from an exponential distribution with $\lambda = 3$. Store the mean of each sample in a vector. Plot a histogram of these sample means. (Be sure to give it a title, and label the axes appropriately.)

```{r}
sim_xbars_exp <- function(m, n, r){
  means <- numeric(m)
  for (i in 1:m) {
    means[i] <- mean(rexp(n, r))
  }
  return(means)
}

hist(sim_xbars_exp(25000, 50, 3), main = "Histogram of Sample Means", xlab = "Sample Means")
```

## Exercise 5 (More Simulation)

Let $X$ follow an exponential distribution with rate parameter $\lambda_X = 2$. Let $Y$ follow a Poisson distribution with rate parameter $\lambda_Y = 3$.

We write $sd(X)$ for the true standard deviation of $X$ and $m(Y)$ for the true median of $Y$.

Let $s_x$ be the sample standard deviation of $X$ which is an estimate of $sd(X)$. Also let $m_y$ be the sample median which is an estimate of $m(Y)$.

Suppose we take samples of size $n_x = 10$ from $X$ and take samples of size $n_y = 5$. Consider the statistic

\[
\frac{s_x}{m_y}.
\]

What is the (sampling) distribution of $\frac{s_x}{m_y}$? We could ask a statistician who specializes in theory. Instead of using mathematics, simulate $\frac{s_x}{m_y}$ 5000 times and store the results. Plot a histogram of the observed values of $\frac{s_x}{m_y}$. Comment on the shape of the histogram and empirical distribution of $\frac{s_x}{m_y}$. Before running your code, set the same seed used for the previous exercise. For full credit, do **not** use a `for` loop.

```{r}
birthday = 19981121
set.seed(birthday)

m <- 5000
sxmy <- numeric(5000)
i <- 0
while (i < m) {
  x <- rexp(10, 2)
  y <- rpois(5, 3)

  sx <- sd(x)
  my <- median(y)
  
  sxmy[i] <- sx/my
  i <- i+1
}
hist(sxmy, main = "Histogram of Sx/My")

```

The shape of the histogram looks like a gamma distribution. 





































